Persistence in functional topology and a correction to a theorem of Morse

### 2021-09-01 func_top.tex
Improve structure of files
Return to original paper size and use biblatex for references

\documentclass{amsart}
\usepackage{mathptmx}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{tikz-cd}
\usepackage[numbers,sort,compress]{natbib}
\usepackage{graphicx}
\usepackage{csquotes}
\usepackage[bookmarks=true,
bookmarksnumbered=true, breaklinks=true,
pdfstartview=FitH, hyperfigures=true,
plainpages=false, naturalnames=true,
colorlinks=true, pagebackref=true,
pdfpagelabels]{hyperref}
\hypersetup{
	linktocpage,
	colorlinks,
	citecolor=blue,
	linkcolor=blue,
	urlcolor=blue}
\usepackage[capitalize, noabbrev]{cleveref}

% layout
\setlength{\textwidth}{\paperwidth}
\addtolength{\textwidth}{-1.85in}
\setlength{\textheight}{\paperheight}
\addtolength{\textheight}{-2in}
\calclayout

% updating to MSC2020
\makeatletter
\@namedef{subjclassname@2020}{%
	\textup{2020} Mathematics Subject Classification}
\makeatother

% theorems
\theoremstyle{plain}
\newtheorem{thm}{Theorem}[section]
\Crefname{thm}{Theorem}{Theorems}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{lem}[thm]{Lemma}
\theoremstyle{definition}
\newtheorem{defi}[thm]{Definition}
\newtheorem{rem}[thm]{Remark}

% commands
\newcommand{\N}{\mathbb{N}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\F}{\mathbb{F}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\HS}{\mathrm{HS}}
\newcommand{\HT}{\mathrm{HT}}
\newcommand{\HLC}{\mathrm{LHS}}
\newcommand{\piT}{\pi\mathrm{T}}
\newcommand{\LC}{\mathrm{LC}}
\newcommand{\NH}{\mathrm{NH}}
\newcommand{\piLC}{\pi\mathrm{LC}}
\newcommand{\HE}{\mathbb{H}^d}
\renewcommand{\H}{H}
\DeclareMathOperator{\interior}{int}
\DeclareMathOperator{\im}{im}
\DeclareMathOperator{\rad}{rad}
\newcommand{\CH}{\check{H}}
\newcommand{\PFD}{\mathrm{PFD}}
\DeclareMathOperator{\Nrv}{Nrv}
\DeclareMathOperator{\Cov}{Cov}
\DeclareMathOperator{\Vietoris}{Vtr}
\DeclareMathOperator{\Balls}{Balls}
\DeclareMathOperator*{\colim}{colim}
\DeclareMathOperator{\crit}{crit}
\newcommand{\mulDom}{\mathcal{E}}
\newcommand{\Top}{\mathsf{Top}}
\newcommand{\Vect}{\mathsf{Vect}}
\renewcommand{\th}{^{\mathrm{th}}}
\newcommand{\m}{\mathfrak{m}}
\newcommand{\n}{\mathfrak{n}}

\begin{document}

@@

	\author[U. Bauer]{Ulrich Bauer}
	\address{U.~Bauer, Technical University of Munich}
	\email{\href{mailto:mail@ulrich-bauer.org}{mail@ulrich-bauer.org}}

	\author[A. Medina-Mardones]{Anibal M. Medina-Mardones}
	\address{A.~Medina-Mardones, Max Planck Institute for Mathematics \and University of Notre Dame}
	\email{\href{mailto:ammedmar@mpim-bonn.mpg.de}{ammedmar@mpim-bonn.mpg.de}}

	\author[M. Schmahl]{Maximilian Schmahl}
	\address{M.~Schmahl, Heidelberg University}
	\email{\href{mailto:mschmahl@mathi.uni-heidelberg.de}{mschmahl@mathi.uni-heidelberg.de}}

	\subjclass[2020]{Primary: 55N31, 58E05. Secondary: 58E12, 54D05, 55N05.}

@@

	\input{abstract}

	\maketitle

	\input{introduction}
	\input{persistence}
	\input{inequalities}
	\input{connectivity}
	\input{surfaces}
	\appendix
	\input{vietoris}

	\input{acknowledgment}

	\bibliographystyle{abbrvnaturl}
	\bibliography{biblio}

\end{document}

>>>>

\documentclass{amsart}
\input{aux/style}
\input{aux/commands}

@@

\author[U.~Bauer]{Ulrich Bauer}
\address{U.B., Technical University of Munich}
\email{\href{mailto:mail@ulrich-bauer.org}{mail@ulrich-bauer.org}}

\author[A.~Medina-Mardones]{Anibal M. Medina-Mardones}
\address{A.M-M., Max Planck Institute for Mathematics \and University of Notre Dame}
\email{\href{mailto:ammedmar@mpim-bonn.mpg.de}{ammedmar@mpim-bonn.mpg.de}}

\author[M.~Schmahl]{Maximilian Schmahl}
\address{M.S., Heidelberg University}
\email{\href{mailto:mschmahl@mathi.uni-heidelberg.de}{mschmahl@mathi.uni-heidelberg.de}}

\subjclass[2020]{55N31, 58E05, 58E12, 54D05, 55N05.}

@@

\begin{document}
	\input{aux/abstract}
	\maketitle
	\input{sec/introduction}
	\input{sec/persistence}
	\input{sec/inequalities}
	\input{sec/connectivity}
	\input{sec/surfaces}
	\appendix
	\input{sec/vietoris}
	\input{aux/acknowledgment}
	\printbibliography
\end{document}

### 2021-09-01 sec/connectivity.tex
Return to original paper size and use biblatex for references

Concretely, let $\H = (\H_d)_{d \in \Z} \colon \Top \to \Vect$ be a fixed graded homotopy invariant functor, which we call a \emph{homology theory}.
A triple of spaces $X_{1}, X_{2} \subseteq X$ is said to have a \emph{Mayer-Vietoris sequence} for $\H$ if the inclusion-induced maps assemble to a long exact sequence
\begin{equation*}
\begin{tikzcd}[column sep = small]
\cdots \arrow[r] & \H_{n}(X_{1} \cap X_{2}) \arrow[r] & \H_{n}(X_{1}) \oplus \H_{n}(X_{2}) \arrow[r] & \H_{n}(X) \arrow[r] & \H_{n-1}(X_{1} \cap X_{2}) \arrow[r] & \cdots \ .
\end{tikzcd}
\end{equation*}
We say that $\H$ has the \emph{open} (resp.\@\ \emph{compact}) \emph{Mayer-Vietories property} if there are natural Mayer-Vietoris sequences for all triples $X_{1}, X_{2} \subseteq X$ with $X = X_1 \cup X_2$ and $X_i \subseteq X$ open (resp.\@\ compact Hausdorff).

For the rest of this section, we will assume that $\H$ is a homology theory that has either the open or the compact Mayer-Vietoris property and for which there is $n_0$ such that $\H_{n}$ is 0 for all $n \leq n_0$.
Note that this includes singular homology with field coefficients, which has the open Mayer-Vietoris property (like any homology theory in the sense of the Eilenberg-Steenrod axioms \cite[Section I]{Eilenberg.1952}) , and it also includes \v{C}ech homology with field coefficients, which has the compact Mayer-Vietoris property.

@@

\begin{proof}
	Recall that we are assuming that the underlying homology theory $\H$ has either the open or the compact Mayer-Vietoris property and that there is some $n_0$ such that $\H_{n}$ is 0 for all $n \leq n_0$.
	The statement of the lemma holds for $\HS_{n}$ in place of $\HS$ for any $n \leq n_0$ since $\H_{n}$ induces the zero map.
	We will proceed by induction on $n \geq n_0$ assuming the statement for $\HS_{n-1}$.

@@

	Finally, using Mayer-Vietoris and the induction hypothesis we will show that $\Sigma_{s,t}$ has the second required property, i.e., that it is closed under finite unions.
	So for $i \in \{1, 2\}$ let $V_i \in \Sigma_{s,t}$ with $U_i$ and $s'_i \in (s,\, t)$ such that 
	$\overline{V_i} \subseteq U_i \subseteq L$ 
	and
	$U_{i} \cap f_{\leq s'_i} \to L \cap f_{\leq t}$
	is $\HS_n$.
	Writing $K_i = \overline{V_i}$, we use \cref{l:neighborhood third} to construct compact sets $K'_i$ such that
	\begin{equation*}
	V_i \subseteq K_i \subseteq V'_i \subseteq K'_i \subseteq U_i \subseteq L
	\end{equation*}
	where $V'_i = \interior(K'_i)$.
	We have that the union $V_1 \cup V_2 \subseteq L$ is open, its closure $\overline{V_1 \cup V_2}$ is compact, and we have $\overline{V_1 \cup V_2} \subseteq V'_1 \cup V'_2 \subseteq L$.
	Thus, we obtain $V_1 \cup V_2 \in \Sigma_{s,t}$ if we can show that there is an $s' \in (s,\, t)$ such that the inclusion 
	$\left(V'_1 \cup V'_2 \right) \cap f_{\leq s'} \to L \cap f_{\leq t}$
	is $\HS_n$.
	To do so, we set $s'' = \min_i s'_i$ and choose $s' \in (s,\, s'')$.
	For proving that $\left(V'_1 \cup V'_2 \right) \cap f_{\leq s'} \to L \cap f_{\leq t}$ is $\HS_n$ we now distinguish the two cases where $\H$ has either the open or the compact Mayer-Vietoris property.

	In the open case, notice that for both choices of $i$ the inclusions
	$U_i \cap f_{\leq s''} \to L \cap f_{\leq t}$
	are $\HS_n$.
	Additionally, the inclusion
	$V'_1 \cap V'_2 \cap f_{\leq s'} \to U_1 \cap U_2 \cap f_{\leq s''}$
	is $\HS_{n-1}$ because it factors through the inclusion
	$K'_1 \cap K'_2 \cap f_{\leq s'} \to U_1 \cap U_2 \cap f_{\leq s''}$,
	which is $\HS_{n-1}$ by the induction hypothesis.
	Because the $V_i$ and $V'_i$ are open and because $\H$ has the open Mayer-Vietoris property, we obtain the following commutative diagram satisfying the assumptions of \cref{l:commutative algebra}:
	\begin{equation*}
	\begin{tikzcd}[column sep=small]
	\H_n(L \cap f_{\leq t}) \oplus \H_n(L \cap f_{\leq t}) \arrow[r] &
	\H_n(L \cap f_{\leq t}) & \\
	\H_{n}(U_1 \cap f_{\leq s''}) \oplus \H_n(U_2 \cap f_{\leq s''}) \arrow[r] \arrow[u] & 
	\H_{n}((U_1 \cup U_2) \cap f_{\leq s''}) \arrow[r] \arrow[u] &
	\H_{n-1}(U_1 \cap U_2 \cap f_{\leq s''}) \\ & 
	\H_{n}((V'_1 \cup V'_2) \cap f_{\leq s'}) \arrow[r] \arrow[u] &
	\H_{n-1}(V'_1 \cap V'_2 \cap f_{\leq s'}) \arrow[u].
	\end{tikzcd}
	\end{equation*}
	We conclude that the inclusion 
	$\left(V'_1 \cup V'_2 \right) \cap f_{\leq s'} \to L \cap f_{\leq t}$ 
	is $\HS_n$, which finishes this part of the proof.

	In the compact case, we apply \cref{l:neighborhood third} again to obtain compact sets $K''_i$ such that
	\begin{equation*}
	V_i \subseteq K_i \subseteq V'_i \subseteq K'_i \subseteq V''_i \subseteq K''_i \subseteq U_i \subseteq L
	\end{equation*}
	where $V''_i = \interior(K''_i)$.
	The rest of the proof is then analogous to the previous case:
	We have that for both choices of~$i$ the inclusions
	$K''_i \cap f_{\leq s''} \to L \cap f_{\leq t}$
	are $\HS_n$ because this is true for the corresponding inclusions with $K''_i$ replaced by $U_i$.
	Moreover, the inclusion
	$K'_1 \cap K'_2 \cap f_{\leq s'} \to K''_1 \cap K''_2 \cap f_{\leq s''}$
	is $\HS_{n-1}$ because it factors through the inclusion
	$K'_1 \cap K'_2 \cap f_{\leq s'} \to V''_1 \cap V''_2 \cap f_{\leq s''}$,
	which is $\HS_{n-1}$ by the induction hypothesis.
	Because the $K'_i$ and $K''_i$ as well as the sublevel sets of $f$ are all compact and because $\H$ has the compact Mayer-Vietoris property, we obtain the following commutative diagram satisfying the assumptions of \cref{l:commutative algebra}:
	\begin{equation*}
	\begin{tikzcd}[column sep=small]
	\H_n(L \cap f_{\leq t}) \oplus \H_n(L \cap f_{\leq t}) \arrow[r] &
	\H_n(L \cap f_{\leq t}) & \\
	\H_{n}(K''_1 \cap f_{\leq s''}) \oplus \H_n(K''_2 \cap f_{\leq s''}) \arrow[r] \arrow[u] & 
	\H_{n}((K''_1 \cup K''_2) \cap f_{\leq s''}) \arrow[r] \arrow[u] &
	\H_{n-1}(K''_1 \cap K''_2 \cap f_{\leq s''}) \\ & 
	\H_{n}((K'_1 \cup K'_2) \cap f_{\leq s'}) \arrow[r] \arrow[u] &
	\H_{n-1}(K'_1 \cap K'_2 \cap f_{\leq s'}) \arrow[u].
	\end{tikzcd}
	\end{equation*}
	We conclude that the inclusion 
	$\left(K'_1 \cup K'_2 \right) \cap f_{\leq s'} \to L \cap f_{\leq t}$
	is $\HS_n$, so the same is true for the inclusion
	$\left(V'_1 \cup V'_2 \right) \cap f_{\leq s'} \to L \cap f_{\leq t}$ 
	because it factors through the previous one.
\end{proof}

>>>>

Concretely, let $\H = (\H_d)_{d \in \Z} \colon \Top \to \Vect$ be a fixed graded homotopy invariant functor, which we call a \emph{homology theory}.
A triple of spaces $X_{1}, X_{2} \subseteq X$ is said to have a \emph{Mayer--Vietoris sequence} for $\H$ if the inclusion-induced maps assemble to a long exact sequence
\begin{equation*}
\begin{tikzcd}[column sep = small]
\cdots \arrow[r] &[-10pt] \H_{n+1}(X) \arrow[d] & &[-20pt] & \\
& \H_{n}(X_{1} \cap X_{2}) \arrow[r] &
\H_{n}(X_{1}) \oplus \H_{n}(X_{2}) \arrow[r] &
\H_{n}(X) \arrow[d] \\ & & &
\H_{n-1}(X_{1} \cap X_{2}) \arrow[r] &
\cdots \ .
\end{tikzcd}
\end{equation*}
We say that $\H$ has the \emph{open} (resp.\@\ \emph{compact}) \emph{Mayer--Vietoris property} if there are natural Mayer--Vietoris sequences for all triples $X_{1}, X_{2} \subseteq X$ with $X = X_1 \cup X_2$ and $X_i \subseteq X$ open (resp.\@\ compact Hausdorff).

For the rest of this section, we will assume that $\H$ is a homology theory that has either the open or the compact Mayer--Vietoris property and for which there is $n_0$ such that $\H_{n}$ is 0 for all $n \leq n_0$.
Note that this includes singular homology with field coefficients, which has the open Mayer--Vietoris property (like any homology theory in the sense of the Eilenberg--Steenrod axioms \cite[Section I]{Eilenberg.1952}) , and it also includes \v{C}ech homology with field coefficients, which has the compact Mayer--Vietoris property.

@@

\begin{proof}
	Recall that we are assuming that the underlying homology theory $\H$ has either the open or the compact Mayer--Vietoris property and that there is some $n_0$ such that $\H_{n}$ is 0 for all $n \leq n_0$.
	The statement of the lemma holds for $\HS_{n}$ in place of $\HS$ for any $n \leq n_0$ since $\H_{n}$ induces the zero map.
	We will proceed by induction on $n \geq n_0$ assuming the statement for $\HS_{n-1}$.

@@

	Finally, using Mayer--Vietoris and the induction hypothesis we will show that $\Sigma_{s,t}$ has the second required property, i.e., that it is closed under finite unions.
	So for $i \in \{1, 2\}$ let $V_i \in \Sigma_{s,t}$ with $U_i$ and $s'_i \in (s,\, t)$ such that
	$\overline{V_i} \subseteq U_i \subseteq L$
	and
	$U_{i} \cap f_{\leq s'_i} \to L \cap f_{\leq t}$
	is $\HS_n$.
	Writing $K_i = \overline{V_i}$, we use \cref{l:neighborhood third} to construct compact sets $K'_i$ such that
	\begin{equation*}
	V_i \subseteq K_i \subseteq V'_i \subseteq K'_i \subseteq U_i \subseteq L
	\end{equation*}
	where $V'_i = \interior(K'_i)$.
	We have that the union $V_1 \cup V_2 \subseteq L$ is open, its closure $\overline{V_1 \cup V_2}$ is compact, and we have $\overline{V_1 \cup V_2} \subseteq V'_1 \cup V'_2 \subseteq L$.
	Thus, we obtain $V_1 \cup V_2 \in \Sigma_{s,t}$ if we can show that there is an $s' \in (s,\, t)$ such that the inclusion
	$\left(V'_1 \cup V'_2 \right) \cap f_{\leq s'} \to L \cap f_{\leq t}$
	is $\HS_n$.
	To do so, we set $s'' = \min_i s'_i$ and choose $s' \in (s,\, s'')$.
	For proving that $\left(V'_1 \cup V'_2 \right) \cap f_{\leq s'} \to L \cap f_{\leq t}$ is $\HS_n$ we now distinguish the two cases where $\H$ has either the open or the compact Mayer--Vietoris property.

	In the open case, notice that for both choices of $i$ the inclusions
	$U_i \cap f_{\leq s''} \to L \cap f_{\leq t}$
	are $\HS_n$.
	Additionally, the inclusion
	$V'_1 \cap V'_2 \cap f_{\leq s'} \to U_1 \cap U_2 \cap f_{\leq s''}$
	is $\HS_{n-1}$ because it factors through the inclusion
	$K'_1 \cap K'_2 \cap f_{\leq s'} \to U_1 \cap U_2 \cap f_{\leq s''}$,
	which is $\HS_{n-1}$ by the induction hypothesis.
	Because the $V_i$ and $V'_i$ are open and because $\H$ has the open Mayer--Vietoris property, we obtain the following commutative diagram satisfying the assumptions of \cref{l:commutative algebra}:
	\begin{center}
		\adjustbox{scale=.92}{
			\begin{tikzcd}[column sep=small]
			\H_n(L \cap f_{\leq t}) \oplus \H_n(L \cap f_{\leq t}) \arrow[r] &
			\H_n(L \cap f_{\leq t}) & \\
			\H_{n}(U_1 \cap f_{\leq s''}) \oplus \H_n(U_2 \cap f_{\leq s''}) \arrow[r] \arrow[u] &
			\H_{n}((U_1 \cup U_2) \cap f_{\leq s''}) \arrow[r] \arrow[u] &
			\H_{n-1}(U_1 \cap U_2 \cap f_{\leq s''}) \\ &
			\H_{n}((V'_1 \cup V'_2) \cap f_{\leq s'}) \arrow[r] \arrow[u] &
			\H_{n-1}(V'_1 \cap V'_2 \cap f_{\leq s'}) \arrow[u].
			\end{tikzcd}
		}
	\end{center}
	We conclude that the inclusion
	$\left(V'_1 \cup V'_2 \right) \cap f_{\leq s'} \to L \cap f_{\leq t}$
	is $\HS_n$, which finishes this part of the proof.

	In the compact case, we apply \cref{l:neighborhood third} again to obtain compact sets $K''_i$ such that
	\begin{equation*}
	V_i \subseteq K_i \subseteq V'_i \subseteq K'_i \subseteq V''_i \subseteq K''_i \subseteq U_i \subseteq L
	\end{equation*}
	where $V''_i = \interior(K''_i)$.
	The rest of the proof is then analogous to the previous case:
	We have that for both choices of~$i$ the inclusions
	$K''_i \cap f_{\leq s''} \to L \cap f_{\leq t}$
	are $\HS_n$ because this is true for the corresponding inclusions with $K''_i$ replaced by $U_i$.
	Moreover, the inclusion
	$K'_1 \cap K'_2 \cap f_{\leq s'} \to K''_1 \cap K''_2 \cap f_{\leq s''}$
	is $\HS_{n-1}$ because it factors through the inclusion
	$K'_1 \cap K'_2 \cap f_{\leq s'} \to V''_1 \cap V''_2 \cap f_{\leq s''}$,
	which is $\HS_{n-1}$ by the induction hypothesis.
	Because the $K'_i$ and $K''_i$ as well as the sublevel sets of $f$ are all compact and because $\H$ has the compact Mayer--Vietoris property, we obtain the following commutative diagram satisfying the assumptions of \cref{l:commutative algebra}:
	\begin{center}
		\adjustbox{scale=.885}{
			\begin{tikzcd}[column sep=small]
			\H_n(L \cap f_{\leq t}) \oplus \H_n(L \cap f_{\leq t}) \arrow[r] &
			\H_n(L \cap f_{\leq t}) & \\
			\H_{n}(K''_1 \cap f_{\leq s''}) \oplus \H_n(K''_2 \cap f_{\leq s''}) \arrow[r] \arrow[u] &
			\H_{n}((K''_1 \cup K''_2) \cap f_{\leq s''}) \arrow[r] \arrow[u] &
			\H_{n-1}(K''_1 \cap K''_2 \cap f_{\leq s''}) \\ &
			\H_{n}((K'_1 \cup K'_2) \cap f_{\leq s'}) \arrow[r] \arrow[u] &
			\H_{n-1}(K'_1 \cap K'_2 \cap f_{\leq s'}) \arrow[u].
			\end{tikzcd}
		}
	\end{center}
	We conclude that the inclusion
	$\left(K'_1 \cup K'_2 \right) \cap f_{\leq s'} \to L \cap f_{\leq t}$
	is $\HS_n$, so the same is true for the inclusion
	$\left(V'_1 \cup V'_2 \right) \cap f_{\leq s'} \to L \cap f_{\leq t}$
	because it factors through the previous one.
\end{proof}

### 2021-09-01 sec/inequalities.tex
Return to original paper size and use biblatex for references

Recall that for a Morse function $f$ on a closed smooth manifold $X$, the usual Morse inequalities state that for any non-negative integer $n$ the following holds:
\begin{equation} \label{e:morse_inequalities}
\sum_{d=0}^n \ (-1)^{n-d} \big( \# \crit_{d}(f) - \beta_{d}(X) \big) \ \geq \ 0,
\end{equation}
where $\crit_{d}(f)$ is the set of critical points of $f$ with index $d$ and $\beta_{d}(X)$ is the $d\th$ Betti number of $X$.

@@

This approach of counting homological changes instead of critical points is also what Morse used in the non-smooth setting of functional topology.
To keep track of the number of $d$-dimensional homological events at filtration value $t$ that persist for at least time $\epsilon > 0$ but not indefinitely, \citet{Morse.1940} defined the $(d, t, \epsilon)$-\emph{cap numbers} of a filtration, which, when the associated persistence module has a barcode and is continuous from above, correspond to the number of finite bars with length greater than $\epsilon$ in the $d\th$ barcode with left endpoint $t$, plus the number of finite bars with length greater than $\epsilon$ in the $(d-1)\th$ barcode with right endpoint $t$ (killed by something $d$-dimensional).

@@

Whenever the sums below are well-defined, we also consider the \emph{$(d,\epsilon)$-cap numbers} 
\begin{equation} \label{e:cap numbers}
m_{d}^{\epsilon} =
\sum_{t} m_{d}^{\epsilon}(t) \ =
\sum_{\substack{(p,q) \in \mulDom \\ q - p > \epsilon \\ q \neq \infty}} \m_{d-1}(p,q) \ +
\sum_{\substack{(p,q) \in \mulDom \\ q - p > \epsilon \\ p \neq -\infty}}\m_d(p,q).
\end{equation}
An important setting where all cap numbers are well-defined is when $M$ is the persistent homology of the sublevel set filtration of a bounded function.
A more general statement can be made using the following notion.

@@

Comparing to the usual Morse inequalities, the cap numbers in dimension $d$ act like the number of critical points with index $d$.
As an analogue to the Betti numbers of the manifold appearing in the usual Morse inequalities, Morse defines quantities $p_{d}$ referred to as \emph{essential dimensions}, which, under the same assumptions as before, can be expressed in the language of persistence diagrams as
\begin{equation} \label{e:essential dimension}
p_{d} \ = \!\! \sum_{p \in \R \cup \{-\infty\}} \m_d(p,\infty),
\end{equation}
which is also the dimension of the colimit of the degree $d$ part of $M$.

@@

\begin{proof}
	Using a telescopic sum argument, the claimed inequalities can be seen to be equivalent to the existence of a sequence $(\nu_d)_d$ of non-negative integers with $m^{\epsilon}_{d} - p_{d} = \nu_{d-1} + \nu_{d}$.
	Since we assume $\m_d(-\infty, p) = 0$ for all $p$, we have
	\begin{align*}
	m^{\epsilon}_{d} \ &=
	\sum_{\substack{(p,q) \in \mulDom \\ q-p > \epsilon \\ q \neq \infty}} \m_{d-1}(p,q) \ +
	\sum_{\substack{(p,q) \in \mulDom \\ q-p > \epsilon \\ p \neq -\infty}} \m_d(p,q) \ =
	\sum_{\substack{(p,q) \in \R^2 \\ q-p > \epsilon }} \m_{d-1}(p,q) \ +
	\sum_{\substack{(p,q) \in \mulDom \\ q-p > \epsilon \\ p \neq -\infty}} \m_d(p,q)
	\end{align*}
	and
	\begin{equation*}
	p_{d} \ = \!\!\!
	\sum_{p \in \R \cup \{-\infty\}} \m_d(p,\infty) \ = \
	\sum_{p \in \R} \m_d(p,\infty),
	\end{equation*}
	which yields
	\begin{align*}
	m^{\epsilon}_{d} - p_{d} \ &= 
	\sum_{\substack{(p,q) \in \R^2 \\ q-p > \epsilon}} \m_{d-1}(p,q) \ +
	\sum_{\substack{(p,q) \in \mulDom \\ q-p > \epsilon \\ p \neq -\infty}} \m_d(p,q) \ - \
	\sum_{p \in \R} \m_d(p,\infty) \ =
	\sum_{\substack{(p,q) \in \R^2 \\ q-p > \epsilon}} \m_{d-1}(p,q)
	\ + \!
	\sum_{\substack{(p,q) \in \R^2 \\ q-p > \epsilon}} \m_d(p,q),
	\end{align*}
	so the sequence given by $\nu_d = \sum_{\substack{(p,q) \in \R^2 \\ q-p > \epsilon}} \m_d(p,q)$ satisfies the equation $m^{\epsilon}_{d} - p_{d} = \nu_{d-1} + \nu_{d}$ as desired.
\end{proof}

As mentioned above, the finiteness assumptions are satisfied if $M$ is initially and eventually constant.
Hence, as a special case, the theorem yields generalized Morse inequalities for any bounded real-valued function whose sublevel set filtration has q-tame persistent homology, including classical Morse functions $f$ on closed smooth manifolds $X$.
As outlined in our motivation for the definition of cap numbers, in this setting our inequalities \eqref{e:morse inequalities} agree with the classical inequalities \eqref{e:morse_inequalities}, as $\beta_d(X) = p_d$ and $m_d^{\epsilon} = \crit_d(f)$ for $\epsilon > 0$ smaller than the minimal difference between any two critical values of~$f$.

>>>>

Recall that for a Morse function $f$ on a closed smooth manifold $X$, the usual Morse inequalities state that for any non-negative integer $n$ the following holds:
\begin{equation} \label{e:classical morse inequalities}
\sum_{d=0}^n \ (-1)^{n-d} \big( \# \crit_{d}(f) - \beta_{d}(X) \big) \ \geq \ 0,
\end{equation}
where $\crit_{d}(f)$ is the set of critical points of $f$ with index $d$ and $\beta_{d}(X)$ is the $d\th$ Betti number of $X$.

@@

This approach of counting homological changes instead of critical points is also what Morse used in the non-smooth setting of functional topology.
To keep track of the number of $d$-dimensional homological events at filtration value $t$ that persist for at least time $\epsilon > 0$ but not indefinitely, \cite{Morse.1940} defined the $(d, t, \epsilon)$-\emph{cap numbers} of a filtration, which, when the associated persistence module has a barcode and is continuous from above, correspond to the number of finite bars with length greater than $\epsilon$ in the $d\th$ barcode with left endpoint $t$, plus the number of finite bars with length greater than $\epsilon$ in the $(d-1)\th$ barcode with right endpoint $t$ (killed by something $d$-dimensional).

@@

Whenever the sums below are well-defined, we also consider the \emph{$(d,\epsilon)$-cap numbers}
\[
m_{d}^{\epsilon} =
\sum_{t} m_{d}^{\epsilon}(t) \ =
\sum_{\substack{(p,q) \in \mulDom \\ q - p > \epsilon \\ q \neq \infty}} \m_{d-1}(p,q) \ +
\sum_{\substack{(p,q) \in \mulDom \\ q - p > \epsilon \\ p \neq -\infty}}\m_d(p,q).
\]
An important setting where all cap numbers are well-defined is when $M$ is the persistent homology of the sublevel set filtration of a bounded function.
A more general statement can be made using the following notion.

@@

Comparing to the usual Morse inequalities, the cap numbers in dimension $d$ act like the number of critical points with index $d$.
As an analogue to the Betti numbers of the manifold appearing in the usual Morse inequalities, Morse defines quantities $p_{d}$ referred to as \emph{essential dimensions}, which, under the same assumptions as before, can be expressed in the language of persistence diagrams as
\[
p_{d} \ = \!\! \sum_{p \in \R \cup \{-\infty\}} \m_d(p,\infty),
\]
which is also the dimension of the colimit of the degree $d$ part of $M$.

@@

\begin{proof}
	Using a telescopic sum argument, the claimed inequalities can be seen to be equivalent to the existence of a sequence $(\nu_d)_d$ of non-negative integers with $m^{\epsilon}_{d} - p_{d} = \nu_{d-1} + \nu_{d}$.
	Since we assume $\m_d(-\infty, p) = 0$ for all $p$, we have
	\begin{align*}
	m^{\epsilon}_{d} \ &=
	\sum_{\substack{(p,q) \in \mulDom \\ q-p > \epsilon \\ q \neq \infty}} \m_{d-1}(p,q) \ +
	\sum_{\substack{(p,q) \in \mulDom \\ q-p > \epsilon \\ p \neq -\infty}} \m_d(p,q) \ =
	\sum_{\substack{(p,q) \in \R^2 \\ q-p > \epsilon }} \m_{d-1}(p,q) \ +
	\sum_{\substack{(p,q) \in \mulDom \\ q-p > \epsilon \\ p \neq -\infty}} \m_d(p,q)
	\end{align*}
	and
	\begin{equation*}
	p_{d} \ = \!\!\!
	\sum_{p \in \R \cup \{-\infty\}} \m_d(p,\infty) \ = \
	\sum_{p \in \R} \m_d(p,\infty),
	\end{equation*}
	which yields
	\begin{align*}
	m^{\epsilon}_{d} - p_{d} \ &=
	\sum_{\substack{(p,q) \in \R^2 \\ q-p > \epsilon}} \m_{d-1}(p,q) \ +
	\sum_{\substack{(p,q) \in \mulDom \\ q-p > \epsilon \\ p \neq -\infty}} \m_d(p,q) \ - \
	\sum_{p \in \R} \m_d(p,\infty) \\ &=
	\sum_{\substack{(p,q) \in \R^2 \\ q-p > \epsilon}} \m_{d-1}(p,q)
	\ + \!
	\sum_{\substack{(p,q) \in \R^2 \\ q-p > \epsilon}} \m_d(p,q),
	\end{align*}
	so the sequence given by $\nu_d = \sum_{\substack{(p,q) \in \R^2 \\ q-p > \epsilon}} \m_d(p,q)$ satisfies the equation $m^{\epsilon}_{d} - p_{d} = \nu_{d-1} + \nu_{d}$ as desired.
\end{proof}

As mentioned above, the finiteness assumptions are satisfied if $M$ is initially and eventually constant.
Hence, as a special case, the theorem yields generalized Morse inequalities for any bounded real-valued function whose sublevel set filtration has q-tame persistent homology, including classical Morse functions $f$ on closed smooth manifolds $X$.
As outlined in our motivation for the definition of cap numbers, in this setting our inequalities \eqref{e:morse inequalities} agree with the classical inequalities \eqref{e:classical morse inequalities}, as $\beta_d(X) = p_d$ and $m_d^{\epsilon} = \crit_d(f)$ for $\epsilon > 0$ smaller than the minimal difference between any two critical values of~$f$.
